Computational methods in landscape ecology

### 2024-10-18 Manuscript.qmd
use modeling spelling

This review provides an overview of recent developments in computational methods that have advanced the research field of landscape ecology.
We focus on key topics such as spatial patterns, connectivity, simulations and modelling, landscape genetics, and spatial planning.

@@

\noindent _Unnithan Kumar et al. (2022) Moving beyond landscape resistance: Considerations for the future of connectivity modelling and conservation science. Landscape Ecology 37:2465–2480._ The paper provides a historical overview of connectivity modelling with a focus on resistance surfaces and discusses current missing aspects and how to address them.

@@

\noindent _Synes et al. (2016) Emerging opportunities for landscape ecological modelling. Current Landscape Ecology Reports 1:146–167._ The paper identifies existing gaps in landscape ecological modelling while highlighting potential emerging opportunities.

>>>>

This review provides an overview of recent developments in computational methods that have advanced the research field of landscape ecology.
We focus on key topics such as spatial patterns, connectivity, simulations and modeling, landscape genetics, and spatial planning.

@@

\noindent _Unnithan Kumar et al. (2022) Moving beyond landscape resistance: Considerations for the future of connectivity modeling and conservation science. Landscape Ecology 37:2465–2480._ The paper provides a historical overview of connectivity modeling with a focus on resistance surfaces and discusses current missing aspects and how to address them.

@@

\noindent _Synes et al. (2016) Emerging opportunities for landscape ecological modeling. Current Landscape Ecology Reports 1:146–167._ The paper identifies existing gaps in landscape ecological modeling while highlighting potential emerging opportunities.

### 2024-10-18 Sections/01-introduction.qmd
use modeling spelling

Landscapes are typically defined in landscape ecology as mosaics of different ecosystems, habitats, land covers, or land use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include, but are not limited to, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change, and human activities within landscapes [@Wu2013KeyConcepts].

>>>>

Landscapes are typically defined in landscape ecology as mosaics of different ecosystems, habitats, land covers, or land use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include, but are not limited to, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modeling, conservation, drivers and consequences of land use and land cover (LULC) change, and human activities within landscapes [@Wu2013KeyConcepts].

### 2024-10-18 Sections/04-connectivity.qmd
use modeling spelling

<!-- R!: Least cost path is just the result of a mathematically formulated problem, so I don't think that it assumes anything. Assumptions are rather in how such results are used in an ecological context -->
Least-cost modelling identifies potential pathways between two points that minimizes the related movement or dispersal costs based on the resistance surface [@Adriaensen2003ApplicationLeastcost].
By calculating pathways between two points it is possible to estimate connectivity between these points based on the accumulated cost along the path.
This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations helping to identify reachable areas within a threshold.
The creation of least-cost paths is a well established technique and there are highly optimized and efficient algorithms available for its calculation.

@@

However, resistance surfaces are not free of critique including missing spatio-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments have shifted towards a combination of both resistance and processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or general computational improvements [@Leonard2017GlowSoftware].

@@

Recently, a new generation of graph-theoretical metrics, commonly reported in combination [@Hashemi2022ReviewEcological], have been developed to account for landscape features that are critical for space use and species-specific movement or dispersal [@Petsas2021LandscapeConnectivity].
Furthermore, the modelling framework was initially only applied to single species, but is increasingly being used to model multispecies landscapes to account for interspecific movement and dispersal abilities and habitat preferences [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach].
In addition, modern multiple-layer graph approaches enable the modelling of not only spatial but also spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
Modelling the temporal dynamics helps to identify habitat patches that impact connectivity over time and the corresponding effects on biodiversity patterns [@Uroy2021HowAssess].

>>>>

<!-- R!: Least cost path is just the result of a mathematically formulated problem, so I don't think that it assumes anything. Assumptions are rather in how such results are used in an ecological context -->
Least-cost modeling identifies potential pathways between two points that minimizes the related movement or dispersal costs based on the resistance surface [@Adriaensen2003ApplicationLeastcost].
By calculating pathways between two points it is possible to estimate connectivity between these points based on the accumulated cost along the path.
This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations helping to identify reachable areas within a threshold.
The creation of least-cost paths is a well established technique and there are highly optimized and efficient algorithms available for its calculation.

@@

However, resistance surfaces are not free of critique including missing spatio-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments have shifted towards a combination of both resistance and processes-based modeling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modeling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or general computational improvements [@Leonard2017GlowSoftware].

@@

Recently, a new generation of graph-theoretical metrics, commonly reported in combination [@Hashemi2022ReviewEcological], have been developed to account for landscape features that are critical for space use and species-specific movement or dispersal [@Petsas2021LandscapeConnectivity].
Furthermore, the modeling framework was initially only applied to single species, but is increasingly being used to model multispecies landscapes to account for interspecific movement and dispersal abilities and habitat preferences [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach].
In addition, modern multiple-layer graph approaches enable the modeling of not only spatial but also spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
Modeling the temporal dynamics helps to identify habitat patches that impact connectivity over time and the corresponding effects on biodiversity patterns [@Uroy2021HowAssess].

### 2024-10-18 Sections/05-sampling.qmd
use modeling spelling

Directly related to the location and extent of sampling sites is the issue of overlapping landscapes which can result in a lower range of landscape heterogeneity, lack of statistical independence, or pseudo-replications [@Hurlbert1984PseudoreplicationDesign; @Zuckerberg2020ReviewOverlapping].
Nonetheless, the lack of independence between sites is likely more related to spatial autocorrelation and thus not necessarily prevented by using non-overlapping landscapes [@Zuckerberg2020ReviewOverlapping].
Instead, spatial autocorrelation between sampling sites can be evaluated using comparisons of similarity measures between sampling approaches or spatial scales (e.g., Moran's I) [@Pasher2013OptimizingLandscape].
Furthermore, spatial autocorrelation in model residuals can be diagnosed using the similarity between two sites as a function of the distance between them, i.e., correlograms [@Dormann2007MethodsAccount].
If spatial autocorrelation is detected researchers may either consider further data collection or use modelling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [@Dormann2007MethodsAccount] or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].

<!--
Software to perform multi-scale analysis include the R package *multilandr*, or the Python libraries *ZonalMetrics* [@Adamczyk2017ZonalMetricsPython], and *PyLandStats*.
-->

>>>>

Directly related to the location and extent of sampling sites is the issue of overlapping landscapes which can result in a lower range of landscape heterogeneity, lack of statistical independence, or pseudo-replications [@Hurlbert1984PseudoreplicationDesign; @Zuckerberg2020ReviewOverlapping].
Nonetheless, the lack of independence between sites is likely more related to spatial autocorrelation and thus not necessarily prevented by using non-overlapping landscapes [@Zuckerberg2020ReviewOverlapping].
Instead, spatial autocorrelation between sampling sites can be evaluated using comparisons of similarity measures between sampling approaches or spatial scales (e.g., Moran's I) [@Pasher2013OptimizingLandscape].
Furthermore, spatial autocorrelation in model residuals can be diagnosed using the similarity between two sites as a function of the distance between them, i.e., correlograms [@Dormann2007MethodsAccount].
If spatial autocorrelation is detected researchers may either consider further data collection or use modeling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [@Dormann2007MethodsAccount] or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].

### 2024-10-18 Sections/06-simulations.qmd
use modeling spelling

Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behavior.
Patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
In landscape ecology individual-based models are increasingly used to model social-ecological systems [@Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [@Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [@Day2020IndividualbasedModeling].
In order to facilitate future model developments recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
Furthermore, hybrid approaches, coupling or integrating different model types with individual-based models, could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].

>>>>

Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behavior.
Patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
In landscape ecology individual-based models are increasingly used to model social-ecological systems [@Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [@Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [@Day2020IndividualbasedModeling].
In order to facilitate future model developments recent progress includes adaptable modeling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
Furthermore, hybrid approaches, coupling or integrating different model types with individual-based models, could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].

### 2024-10-18 Sections/09-conclusion.qmd
use modeling spelling

Since its emergence in the 1980s [@With2019EssentialsLandscape], landscape ecology research has constantly evolved and is now highly dependent on computational methods.
Thus, future progress within the field will largely depend on available software and computational tools [@Costanza2019DescribingAnalyzing].
Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modelling, landscape genetics and spatial planning.
In this context, we highlight open-source software as a cornerstone of "Open Science" offering key advantages like shareability, reproducibility, and transparency, which provide great benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].

@@

Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
This includes advances in remote sensing technology with increasing data availability and diversity from a range of sources, including satellites, unmanned aerial vehicles, and ground-based sensor systems.
These systems provide both passive sensors, such as multispectral, hyperspectral, and thermal sensors and active optical sensors, such as LiDAR and SAR [@Crowley2020RemoteSensings].
Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible [@Markham2023ReviewMethods].
In combination with other increasingly available data sources, e.g, citizen science data, ecology has entered the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
Due to its high flexibility and performance artificial intelligence has become popular in ecology, and deep learning and machine learning algorithms can be used for, e.g., mapping, classifying and extracting features, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, and documentation [@Tenopir2020DataSharing].

>>>>

Since its emergence in the 1980s [@With2019EssentialsLandscape], landscape ecology research has constantly evolved and is now highly dependent on computational methods.
Thus, future progress within the field will largely depend on available software and computational tools [@Costanza2019DescribingAnalyzing].
Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modeling, landscape genetics and spatial planning.
In this context, we highlight open-source software as a cornerstone of "Open Science" offering key advantages like shareability, reproducibility, and transparency, which provide great benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].

@@

Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
This includes advances in remote sensing technology with increasing data availability and diversity from a range of sources, including satellites, unmanned aerial vehicles, and ground-based sensor systems.
These systems provide both passive sensors, such as multispectral, hyperspectral, and thermal sensors and active optical sensors, such as LiDAR and SAR [@Crowley2020RemoteSensings].
Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible [@Markham2023ReviewMethods].
In combination with other increasingly available data sources, e.g, citizen science data, ecology has entered the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
Due to its high flexibility and performance artificial intelligence has become popular in ecology, and deep learning and machine learning algorithms can be used for, e.g., mapping, classifying and extracting features, modeling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, and documentation [@Tenopir2020DataSharing].
